\documentclass{llncs}
\pagestyle{plain} % turn on page numbers in llncs

\usepackage{isabelle,isabellesym}
\usepackage{latexsym}
\usepackage{amssymb}
\usepackage{mathtools}

% this should be the last package used
\usepackage{pdfsetup}

\newtheorem{thm}{Theorem}

% no right margin in quote:
\renewenvironment{quote}
{\list{}{}%
\item\relax}
{\endlist}

\newcommand{\noquotes}[1]{{\renewcommand{\isachardoublequote}{}\renewcommand{\isachardoublequoteopen}{}\renewcommand{\isachardoublequoteclose}{}#1}}

\isabellestyle{it}

\renewcommand{\isacharunderscore}{\_}
\renewcommand{\isacharunderscorekeyword}{\_}
\renewcommand{\isadigit}[1]{{\rm #1}}

% for uniform font size
\renewcommand{\isastyle}{\isastyleminor}

\newcommand{\eqnum}{\refstepcounter{equation}\hfill(\theequation)}

\hyphenation{Isa-belle}


\begin{document}

\title{Verification of a Closest Pair of Points Algorithm}
\author{Martin Rau and Tobias Nipkow}
\institute{Technische Universit\"at M\"unchen}
\date{}
\maketitle

\begin{abstract}
We verify a divide-and-conquer algorithm, solving one of the fundamental problems
in \textit{Computational Geometry}, the \textit{Closest Pair of Points} problem.
Using the interactive theorem prover Isabelle/HOL, we prove functional correctness
and the optimal running time of $\mathcal{O}(n \log n)$ of the algorithm. 
Additionally we generate executable code which is empirically competitive with handwritten
 reference implementations.
\end{abstract}


\section{TODO}

\begin{itemize}
\item Add link to AFP entry in introduction \\
\item Local copy of LatexSugar \\
\item Adjust set conversions \\
\item Log formatting \\log \\
\item General formatting \\
\item Write Isabelle notation subsection in introduction \\
\item Write running time section \\
\item Write executable code section \\
\item Write related work section \\
\item Update benchmarks \\
\item Write AFP entry \\
\end{itemize}


\section{Introduction}

The \textit{Closest Pair of Points} or \textit{Closest Pair} problem is one of the fundamental
problems in \textit{Computational Geometry}: Given a set $P$ of $n \geq 2$ points in $\mathbb{R}^d$,
find the closest pair of $P$, i.e. two points $p_0 \in P$ and $p_1 \in P$ ($p_0 \ne p_1$) such that 
the distance between $p_0$ and $p_1$ is smaller than or equal to the distance of any distinct pair 
of points of $P$.

Shamos and Hoey \cite{Closest-Point-Problems:1975} are one of the first to mention this problem and
introduce an algorithm based on \textit{Voronoi} diagrams for the planar case, improving the running 
time of the best known algorithms at the time from $\mathcal{O}(n^2)$ to
$\mathcal{O}(n\ \mathit{log}\;n)$. They also prove that this running time is optimal for a
deterministic computation model. One year later, in 1976, Bentley and Shamos
\cite{Divide-And-Conquer-In-Multidimensional-Space:1976} publish a, also optimal, divide-and-conquer
algorithm to solve the Closest Pair problem that can be non-trivially extended to work on
spaces of arbitrary dimension. Since then the problem has been the focus of extensive research and
a multitude of optimal algorithms have been published. Smid \cite{Handbook-Computational-Geometry:2000}
provides a comprehensive overview over the available algorithms, including randomized approaches which
improve the running time even further to $\mathcal{O}(n)$.

The main contributions of this paper are the verification of a functional implementation of a
divide-and-conquer algorithm solving the Closest Pair problem for the two-dimensional Euclidean plane
with the optimal running time of $\mathcal{O}(n\ \mathit{log}\;n)$. We use the interactive theorem 
prover Isabelle/HOL \cite{LNCS2283,Concrete} to prove functional correctness as well as the 
running time of the algorithm. Empirical testing also shows that our verified algorithm is 
competitive with handwritten reference implementations. The basis for our implementation 
is the algorithm presented by Cormen et al. \cite{Introduction-to-Algorithms:2009}.

This paper is structures as follows:
Section \ref{section:closest_pair_algorithm} familiarizes the reader with the algorithm by presenting a
high-level description. Section \ref{section:proving_functional_correctness} presents the algorithm
implementation and the functional correctness proof. Section \ref{section:proving_running_time} introduces
the reader to our methodology of formalizing running time proofs in Isabelle/HOL and proceeds to prove 
the running time of $\mathcal{O}(n\ \mathit{log}\;n)$. Section \ref{section:executable_code}
describes final adjustments to obtain an executable version of the algorithm in target languages
such as OCaml and SML and evaluates it against handwritten imperative and functional implementations. 
Section \ref{section:conclusion} concludes with summarizing our results and listing related and future work.

\input{Paper.tex}

\section{Conclusion} \label{section:conclusion}

We did it.

\paragraph{Acknowledgement}
XYZ helped.

\bibliographystyle{splncs04}
\bibliography{root}

\end{document}
